Material Culture Networks in Tonto Basin

### 2022-12-28 manuscript.tex
finished updating paper based on Chris Caseldine's comments

---
title: Material Culture Networks in Tonto Basin
output:
  rticles::tf_article: null
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities. This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
always_allow_html: true
---

```{r setup, echo = F}
out.format <- knitr::opts_knit$get("out.format")
# options(knitr.table.format = "latex")
knitr::opts_chunk$set(
  echo = F,
  message = FALSE,
  warning = FALSE,
  dpi = 450,
  out.width = "100%"
)
```

```{r load, include=FALSE}
library(tidyverse,quietly = T)
library(magrittr)
library(Momocs,quietly = T)
library(kableExtra)
library(huxtable)
library(ggthemes, quietly = T)
library(here)
library(rio)
library(igraph)
library(ggnetwork)
library(vegan)
library(multinet)

set.seed(1010)

# library(ggimage)
i_am("manuscript.Rmd")
# keep track of figures and tables
figs = 1
tbls = 1
set.seed(1010)

# load data and remove sites with only one projectile point

sites = import(here("data/TontoBasinNetworkSites.xlsx"),setclass = 'tibble') %>% 
  ungroup() %>% 
  filter(!ArchaMapName %in% c("AZ V:5:130 (ASM)","AZ U:4:009 (ASM)","AZ U:8:318 (ASM)","AZ V:5:121 (ASM)"))

# data frame for network nodes
nodes = sites %>%
  select(ArchaMapName,Area,Architecture = architecture) %>% 
  mutate(Architecture = case_when(str_detect(Architecture,"compound, platform mound")~"platform mound",str_detect(Architecture,"compound, roomblock")~"roomblock",TRUE~Architecture)) %>% 
  rowid_to_column("id") %>% 
  ungroup()

# coords = outTonto$fac %>% 
#   select(ArchaMapName,UTMEAST,UTMNORTH) %>% 
#   distinct_all()
# coords %>% export(here('data/siteCoords.xlsx'))
coords = import(here('data/siteCoords.xlsx')) %>% 
  inner_join(nodes) %>% 
  arrange(id)
CeramicData = rio::import(here("data/CeramicData.xlsx"),
                          setclass='tibble')

# number of types per site
count = CeramicData %>% as.data.frame() %>%
  group_by(ArchaMapName,SWSN_type) %>%
  summarize(Count = sum(Count)) %>%
  ungroup()
table(count$ArchaMapName)
ceramicDF = CeramicData %>% as.data.frame() %>%
  group_by(ArchaMapName,SWSN_type) %>%
  summarize(Count = sum(Count)) %>%
  ungroup() %>%
  inner_join(nodes %>% select(ArchaMapName,id)) %>% 
  select(-ArchaMapName) %>% 
  pivot_wider(names_from = SWSN_type, values_from = Count) %>%
  column_to_rownames("id") %>% 
  select(-`Type, Undifferentiated`)

outTonto = readRDS(here("data/TontoNetworkOutlines.Rds"))

outTonto$fac %<>% select(-id,-Area) %>% left_join(nodes)
# outTonto %<>%
# mutate(ArchaMapName = droplevels(ArchaMapName))
# outTonto %>% saveRDS(here("data/TontoNetworkOutlines.Rds"))
```

```{r sessionInfo,eval=FALSE, include=FALSE}
# display software information
sessioninfo::session_info()
```

# Introduction {-}

Network science has many applications for archaeologists. It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society. But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks. Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help archaeologists direct future research. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub]. This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg]. There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites. In reality, I expected significant variation between these types of material culture. As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance. Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences. I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.

## Tonto Basin

Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview]. The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p. 129-133]. The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian. All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.

```{r TontoBasinSites, fig.cap= 'Map of archaeology sites from the Roosevelt Platform Mounds Study in Tonto Basin. Groups use the original assignments.'}

figs = figs + 1
knitr::include_graphics(here("figures/TontoBasinSitesv3.png"))
```

Tonto Basin was occupied, although likely not continuously, from the Archaic through the late Classic period. The height of occupation, particularly the sites excavated during the project, dates between AD 1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt phase is notable for the appearance of Salado pottery and platform mounds--probably introduced from the Phoenix Basin [@Elson1996-ym]. Moderate occupation continued into the Gila phase (AD 1325-1450), which was characterized by population loss and aggregation into larger sites. This phase marks the end of recognizable occupation in the area until the historic period.

One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu]. This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

# Network Analysis {-}

Network analysis is a flexible tool for analyzing many types of data [see @Brughmans2023-uj]. The key element is that some way must be determined to tie each entity together. This analysis uses sites as nodes in the network and some type of similarity as a tie to connect each node. Each type of similarity depends on the type of data. Table `r tbls` lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site. The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used. The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku]. This is not a random sample of Tonto Basin sites. It is merely a sample of convenience. Sampling can have a major impact on conclusions. In this case, I am not drawing conclusions based on a complete network or the existence of a random sample (though biases can still affect the results). Rather, I am interested in whether the types of material culture present in this study co-vary with each other and with space. These questions do not require a complete network.

## Defining Networks and Relations

Network approaches are best suited where the relationships between entities are of primary interest. In this case, the entities are defined as archaeology sites. The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind. Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study. His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin. Unfortunately, this study will only attempt a site-level analysis. Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis. He used similarity networks where the points that had similar flaking styles were connected together. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts. This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.

Yet, it can be difficult to understand how a similarity network is related to a past social network. What kinds of interactions are represented by similarities in projectile points or in ceramics? Answering this question is also a crucial step in interpreting networks. One way to examine relations between individuals who make or use similar types of projectile points or pots is to talk about identity. Identity can be a troubling topic for anthropologists. There are numerous meanings given to it and numerous scales at which it applies [@Brubaker2000-na], but Peeples [-@Peeples2018-ib] has introduced archaeologists to a more practical approach that is useful here. This approach views identity as existing along two axes: one categorical and the other relational. Relational identification is the process of identifying with someone due to frequent interaction. Categorical identification is the process of identifying with someone because you belong to a recognized social group. For example, members of the same moiety would share a categorical identity. They would likely also share a relational identity if they frequently interacted. These identities can be reflected in material culture. Pottery makes a good example. Peeples [-@Peeples2018-ib, p. 151] notes that bowls in the Cibola region were sometimes painted with a bright, red slip with designs painted on the interior and they were among the first in the Southwest to have designs on the exterior. This is a strong indication of categorical identity. The potters were attempting to make a clear signal for whoever saw the pot. Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils. The particulars of these actions would be learned from close interaction with a teacher or other potters. In this way, relational identity can be compared to communities of practice [@Lave1991-ik; @Wenger1998-rx].

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership. What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study. Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity. He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative. Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz]. Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. This is indicative of relational identity. There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone. Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].

There is one other aspect to these relations that I believe played an important role. A central way people identify themselves is by gender. Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but this fits the available data for the Hohokam [@Crown1996-xb; @VanPool2010-im, p. 253]. An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7]. On the other hand, projectile points were almost exclusively buried with males. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.

@@

The network for architecture, ceramics, and projectile points, were created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site. This is a simple measure of similarity that does not take into account abundance. It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
  # theme_article() %>% 
  # set_caption(cap)
  knitr::kable(caption= cap,
               booktabs = T,
               format.args = list(big.mark = ",",
                                  scientific = FALSE)) %>%
  kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

>>>>

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\def\UrlBreaks{\do\/\do-\do?}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\articletype{}

\title{Material Culture Networks in Tonto Basin}

\author{\name{Robert J. Bischoff$^{a}$}
\affil{$^{a}$School of Human Evolution and Social Change, Arizona State
University, Tempe, AZ}
}

\thanks{CONTACT Robert J.
Bischoff. Email: \href{mailto:rbischoff@asu.edu}{\nolinkurl{rbischoff@asu.edu}}}

\maketitle

\begin{abstract}
Network analysis has a strong foundation in Southwest archaeology, yet
the combination of multiple types of artifacts in one
analysis--multilayer network analysis--has not been formally applied
except within a single artifact type. Many studies consider material
culture holistically, yet network analysis has the advantage of focusing
specifically on the relationships between entities (often communities)
and how the structure of a network can benefit particular entities. This
study takes architecture, ceramic, projectile point, and site location
data from the Roosevelt Platform Mound Study and combines these data
into a multilayer network analysis. This analysis provides a way to test
the co-variance of these types of material culture with each other and
with spatial variation. The results demonstrate a strong association
between projectile points and spatial distance. Sites containing
roomblock architecture are highly central in the ceramic network but
have low centrality in the projectile point network. Overall, the
ceramic and projectile point networks exhibit significant differences.
This indicates that the social networks that created these patterns had
different social mechanisms. One potential cause of these differences is
gendered spheres of interaction with men exchanging projectile points
and women exchanging ceramics.
\end{abstract}

\begin{keywords}
American Southwest; network analysis; multilayer networks; Hohokam;
projectile points; ceramics; computational archaeology; gender
\end{keywords}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Network science has many applications for archaeologists. It can be
particularly useful for breaking out of traditional spatial categories
and examining data through new lenses
\citep{Feinman2020-nf, Holland-Lulewicz2021-eb}. Multilayer network
analysis combines multiple networks into one analytical framework
\citep{Kivela2014-mh}. Social systems are complex webs of
interrelations. There is too much involved to model the material culture
of even a technologically simple society. But certainly analyzing more
than one type of material culture provides a more complete understanding
of the social dynamics involved in the networks of interaction we wish
to understand. Yet, there are few examples of multilayer networks in
archaeology \citep{Giomi2022-dn, Upton2019-yg} and none I am aware of
that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there
are examples of network studies that consider multiple types of material
culture. For example, Mills and colleagues \citeyearpar{Mills2013-gc}
used ceramics and obsidian to understand the transformation of large
scale social networks. Peeples \citeyearpar{Peeples2018-ib} used
ceramics and architecture to understand identity and social change in
the Cibola region. This paper is an application of multilayer network
analysis that seeks to further network research by examining multiple
types of material culture in one framework. Given the frequent focus on
single types of artifacts in network research, a primary question I seek
to address is whether networks based on different types of material
culture are positively correlated. This is an important question because
archaeologists often make inferences based solely on ceramics or other
types of material culture. Evidence regarding the ways material culture
co-vary will help archaeologists direct future research. Furthermore, I
also seek causal explanations for the co-variance, or lack thereof, for
material culture in the case study I present here.

Tonto Basin holds great potential for archaeological research due to the
large cultural resource management projects conducted in the region
primarily in the 1980s-1990s
\citep{Ahlstrom1991-eb, Ciolek-Torrello1994-ds, Doelle1992-lu, Rice1998-ku},
as well as the large number of syntheses and other studies focusing on
the area
\citep[e.g.,][]{Clark2004-uw, Dean2000-uc, Elson1995-tb, Elson2000-zq, Hill2015-at, Huntley2016-em, Lange1992-gp, Lyons2003-yy, Lyons2006-to, Lyons2012-pr, Neuzil2008-zd, Oliver2001-tu, Stark1998-mu, Watts2013-ub}.
This analysis uses data from the Roosevelt Platform Mounds Study (RPMS),
the largest of these projects \citep{Rice1998-ku}, to examine basic
architectural data, typed ceramics, and a recent projectile point
analysis \citep{Bischoff2022-rg}. There are two null hypotheses tested
in this analysis: (1) that the data is spatially correlated--meaning
that sites nearest to each other will be most alike--and (2) that the
architecture, ceramics, and projectile points will all be positively
correlated--meaning that similar types of architecture, ceramics, and
projectile points will be found at the same sites. In reality, I
expected significant variation between these types of material culture.
As will be demonstrated, the results indicate a positive correlation
between projectile point similarity and spatial distance. Furthermore,
ceramic and projectile point networks (hereafter point networks) exhibit
significant differences. I posit that the differences in these networks
are connected to the identities of the individuals creating the original
social networks: particularly that of gender.

\hypertarget{tonto-basin}{%
\subsection{Tonto Basin}\label{tonto-basin}}

Tonto Basin is located in east-central Arizona and features Tonto Creek
flowing from the Mogollon Rim into the Salt River (figure 1). Parts of
the region have been intensively studied, while certain time periods and
much of the uplands have been less intensively sampled \citep[see][ for
a recent overview]{Clark2021-lu}. The basin was also conveniently
located along regional travel routes, allowing major
settlements--particularly those with platform mounds--to participate and
benefit from this exchange
\citep[p.~129-133]{Caseldine2022-uu, Wood2000-ze}. The junction of the
Salt River and Tonto Creek now forms Roosevelt Lake since the damming of
the confluence of the Salt River and Tonto Creek. It was the expansion
of the dam that precipitated the RPMS and its related projects in the
late 1980s-1990s. Some important regional differences may have
differentiated sites along the Salt River and Tonto Creek arms of
Roosevelt Lake \citep{Lyons2013-ya, Simon2001-am} based on non-local
ceramics and obsidian. All of the artifacts/data from the RPMS project
are hosted at the Center for Archaeology and Society at Arizona State
University, and the data in this study come exclusively from this
project.

\begin{figure}
\includegraphics[width=1\linewidth]{figures/TontoBasinSitesv3} \caption{Map of archaeology sites from the Roosevelt Platform Mounds Study in Tonto Basin. Groups use the original assignments.}\label{fig:TontoBasinSites}
\end{figure}

Tonto Basin was occupied, although likely not continuously, from the
Archaic through the late Classic period. The height of occupation,
particularly the sites excavated during the project, dates between AD
1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt
phase is notable for the appearance of Salado pottery and platform
mounds--probably introduced from the Phoenix Basin \citep{Elson1996-ym}.
Moderate occupation continued into the Gila phase (AD 1325-1450), which
was characterized by population loss and aggregation into larger sites.
This phase marks the end of recognizable occupation in the area until
the historic period.

One aspect of Tonto Basin that has caught the attention of researchers
is its location in a border zone between the Hohokam, Mogollon, and
Ancestral Pueblo regions
\citep[e.g.,][]{Caseldine2022-uu, Clark2001-ck, Elson1994-hk, Hill2015-at, Huntley2016-em, Lyons2012-pr, Lyons2013-ya, Neuzil2008-zd, Wood2000-ze}.
Of particular note is the presence of masonry roomblock architecture and
pottery uncharacteristic of the local Hohokam traditions, which has been
interpreted as Kayenta immigration into Tonto Basin
\citep{Clark2001-ck, Lyons2003-yy, Lyons2006-to, Stark1998-mu}. This
migration into the Tonto Basinwas part of a larger migration from north
to south and is associated with the origin of the Salado phenomenon.
Salado pottery production was widespread across southern Arizona, but
often the largest sources of production was centered at the location of
a former Kayenta enclave
\citep{Hill2015-at, Huntley2016-em, Lyons2012-pr, Neuzil2008-zd}. The
connection between immigration and an important type of decorated
pottery in the region suggests that sites with roomblocks may have
higher centrality in ceramic networks. Having high centrality means they
have more connections to other nodes (i.e., sites). The presence of
migrant communities in Tonto Basin invites questions regarding the
relationship between sites--particularly between migrant and local
communities. A network approach is an ideal way to examine the
relationships between sites. Evidence that projectile points move
between sites \citep[e.g.,][]{Watts2013-ub} indicates that this is also
a productive line of analysis, and one that is less-commonly pursued
compared to architecture and ceramics.

\hypertarget{network-analysis}{%
\section*{Network Analysis}\label{network-analysis}}
\addcontentsline{toc}{section}{Network Analysis}

Network analysis is a flexible tool for analyzing many types of data
\citep[see][]{Brughmans2023-uj}. The key element is that some way must
be determined to tie each entity together. This analysis uses sites as
nodes in the network and some type of similarity as a tie to connect
each node. Each type of similarity depends on the type of data. Table 1
lists the 11 sites with sufficient data to include in this study with
the types of architecture present, periods occupied, and total number of
ceramics and complete projectile points found at each site. The lack of
complete projectile points at some sites is the largest caveat in this
study, but complete points were necessary for the type of projectile
point analysis used. The periods, architecture types, and ceramic types
were assigned during the RPMS project and can be found in the project
reports \citep{Rice1998-ku}. This is not a random sample of Tonto Basin
sites. It is merely a sample of convenience. Sampling can have a major
impact on conclusions. In this case, I am not drawing conclusions based
on a complete network or the existence of a random sample (though biases
can still affect the results). Rather, I am interested in whether the
types of material culture present in this study co-vary with each other
and with space. These questions do not require a complete network.

\hypertarget{defining-networks-and-relations}{%
\subsection{Defining Networks and
Relations}\label{defining-networks-and-relations}}

Network approaches are best suited where the relationships between
entities are of primary interest. In this case, the entities are defined
as archaeology sites. The archaeology sites in turn represent a group of
individuals who lived or visited the area of the site and left material
remains behind. Ideally, studying social relations would be approached
at the individual level, but rarely can archaeologists address
archaeological data with that level of specificity. Watts
\citeyearpar{Watts2013-ub} used projectile points from Tonto Basin in
just such a study. His analysis of flake scar patterning identified
individual knappers, or at least clusters of knappers with similar
knapping styles, and the distribution of their projectile points around
Tonto Basin. He found strong connections between many parts of the
eastern Tonto Basin. Unfortunately, this study will only attempt a
site-level analysis. Watt's study does, however, illustrate how
relations can be defined between nodes in a network analysis. He used
similarity networks where the points that had similar flaking styles
were connected together. Similarity networks are a commonly used type of
network in archaeology
\citep[e.g.,][]{Birch2018-xx, Borck2015-jy, Cochrane2010-oa, Golitko2015-we, Lulewicz2019-lu, Mills2013-wq, Peeples2018-ib, Terrell2010-hd}.
These studies use a variety of artifacts and methods to construct their
networks, but each has demonstrated the utility of network methods
within archaeological contexts. This study also uses similarity networks
to group sites by similarity in ceramic assemblages and similarity in
projectile point forms.

Yet, it can be difficult to understand how a similarity network is
related to a past social network. What kinds of interactions are
represented by similarities in projectile points or in ceramics?
Answering this question is also a crucial step in interpreting networks.
One way to examine relations between individuals who make or use similar
types of projectile points or pots is to talk about identity. Identity
can be a troubling topic for anthropologists. There are numerous
meanings given to it and numerous scales at which it applies
\citep{Brubaker2000-na}, but Peeples \citeyearpar{Peeples2018-ib} has
introduced archaeologists to a more practical approach that is useful
here. This approach views identity as existing along two axes: one
categorical and the other relational. Relational identification is the
process of identifying with someone due to frequent interaction.
Categorical identification is the process of identifying with someone
because you belong to a recognized social group. For example, members of
the same moiety would share a categorical identity. They would likely
also share a relational identity if they frequently interacted. These
identities can be reflected in material culture. Pottery makes a good
example. Peeples \citeyearpar[p.~151]{Peeples2018-ib} notes that bowls
in the Cibola region were sometimes painted with a bright, red slip with
designs painted on the interior and they were among the first in the
Southwest to have designs on the exterior. This is a strong indication
of categorical identity. The potters were attempting to make a clear
signal for whoever saw the pot. Relational identity can be seen in the
way the potter prepares the clay recipe and smooths the coils. The
particulars of these actions would be learned from close interaction
with a teacher or other potters. In this way, relational identity can be
compared to communities of practice \citep{Lave1991-ik, Wenger1998-rx}.

In this analysis, I argue that ceramics and architecture are more likely
to represent categorical identity. The architecture discussed is highly
visible and representative of historical group membership. The ceramics
are grouped into types based primarily on decoration, which is strongly
indicative of group membership. What this means is that links between
nodes in the cer The details of architecture and ceramics used in this
study are discussed in a future section, but it is important to note
that these designations as categorical or relational are contextually
dependent on this study. Clark \citeyearpar{Clark2001-ck} has an
excellent discussion on material culture and its relationship to
identity. He identifies certain patterns, but his synthesis and other
research
\citep{Carr1995-mq, Carr1995-sh, Dietler1998-zo, Gosselain1998-qd, Gosselain2000-wf, Gosselain2016-as, Hodder1982-no, Huntley2008-qu, Lemonnier1986-le, Lyons2003-yy, Neuzil2008-zd, Sassman2001-sw, Stark1998-mu, Wiessner1983-ei, Wiessner1997-su}
strongly indicates that the relationship between material culture and
identity is culturally relative. Projectile points can in some cases
indicate group membership \citep{Wiessner1983-ei, Wiessner1997-su}, but
in this case the triangular and side-notched points were likely used for
different purposes \citep[see][]{Loendorf2015-ww, Sliva2002-oz}. Styles
were difficult to determine for the original researchers
\citep{Rice1994-rk}, and I had the same difficulty. In my opinion, the
subtle differences between projectile point outlines are more
representative of interactions between knappers than markers of group
identity for this particular case. This is indicative of relational
identity. There are no environmental reasons to assume differences in
projectile points, as each site was in a similar ecozone. Each hunter
would have been using the points for the same game and differences in
point styles would have been primarily cultural \citep{Sliva2002-oz}.

There is one other aspect to these relations that I believe played an
important role. A central way people identify themselves is by gender.
Gender roles, like identity, can also vary in complex ways, but for the
purposes of this analysis I will use a simplified model. Women made
pottery and men made projectile points. This was not always true of
course, but this fits the available data for the Hohokam
\citep[p.~253]{Crown1996-xb, VanPool2010-im}. An examination of the
burial record for the RPMS study shows no obvious indication that women
made pottery, only that both men and women were buried with pottery
\citep[Table 10.7]{Loendorf1998-ln}. On the other hand, projectile
points were almost exclusively buried with males. This does not mean
that a man could not move pots from one site to another or that a woman
could not do the same with a projectile point, but, in general,
differences in point and ceramic networks are most likely to indicate
differences in gender networks.

@@

The network for architecture, ceramics, and projectile points, were
created using the Jaccard similarity coefficient \citep{Jaccard1912-na}
to determine the links between each site. This is a simple measure of
similarity that does not take into account abundance. It was chosen due
to the large differences in sample sizes between sites for the ceramics
and projectile points and because count data were irrelevant or not
readily available for the architectural data.

\begin{table}
